---
title: "Statistical Regression Assignment 1"
author: "Moreen Mbuki"
date: "2024-11-12"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A retail company wants to understand the customer purchase behaviour
(specifically, purchase amount).First, we need to load the data set
data_purchase_behaviors which will be used to to understand the customer
purchase behaviour (specifically,purchase amount). The data-set contains
the following columns: User ID, Gender, City Category, Stay In Current
City Years (this is the length of stay in the current city), Marital
Status (coded as 1 for married, 0 otherwise), Age num (this is the
approximate age of the user), and Purchase (this is the purchase
amount).

This data-set contain 159000 observation with 7 variables which are both
numerical data-set: Age_num of the customers, purchase amount, Atay in
the current city in years and categorical data-set which are as follows
: Marital status, Gender and City categories.For the case id descriptive
analysis the data-set is partition into the two types.

```{r, echo=FALSE,cache.comments=TRUE,comment= NA,include=FALSE}
# installing the required packages 
# install.packages("ggplot2")
# install.packages("Cairo")
# install.packages("readr")
# install.packages("psych")


# loading the necessary packages
library(Cairo)
library(ggplot2)
library(readr)
library(stats4)
library(psych)

```

```{r, include=FALSE, echo=FALSE,cache.comments=TRUE}
library(readr)
library(stats4)
setwd("C:/Users/Administrator/Desktop/AIMS/Review Phase/Statisticical Regression/0 - Getting Started with R")
```

```{r, echo=FALSE,cache.comments=TRUE,comment = NA,include=FALSE}
data <- read.csv("data_purchase_behaviour.csv")
head(data,3)

```

```{r,include=FALSE,echo=FALSE,cache.comments=TRUE,comment = NA}
str(data)
```

```{r, echo=FALSE,cache.comments=TRUE,include=FALSE,comment = NA}
colnames(data)
```

```{r, echo=FALSE,cache.comments=TRUE,include=FALSE,comment = NA}
summary(data)
```

To start with is numerical variables, after sub-setting we can start by
have a look at the data set.

```{r, echo=FALSE,cache.comments=TRUE,comment = NA,include=FALSE}
num_data <-data[,c("Stay_In_Current_City_Years","Age_num", "Purchase")]
head(num_data,3)
```

From the descriptive statistics of the data set,the maximum purchase
amount is 23,961 with the minimum of 12, for the age of the customers
the maximum age is 75 years while the minimum age is 10 years. The
customers the maximum stay in the current city being 4.

```{r, echo=FALSE,cache.comments=TRUE,comment = NA,warning=FALSE, include=FALSE}
library(psych)
describe(num_data)
```

Similarly we can use box-plots to clearly see the distribution of the
data-set. The figure below show the subplot for the stay in the current
city in years , Age and the purchase respectively.

```{r, echo=FALSE,cache.comments=TRUE,comment = NA,fig.height= 3 , fig.width=8, out.width='100%', fig.cap="BoxPlot for Age,Puchase and Current City"}
#pdf("box-plot.pdf",width = 12, height = 5)
par(mfrow = c(1,3))
boxplot(num_data$Stay_In_Current_City_Years,
        main = "Boxplot Current_City_Years",
        xlab = "Stay_In_Current_City_Years",
        ylab = "Stay_In_Current_City_Years ",
        col = "pink",  
        border = "black")
boxplot(num_data$Age_num,
        main = "Boxplot of Age_num",
        xlab = "Age_num",
        ylab = "Age ",
        col = "blue",  
        border = "black")

boxplot(num_data$Purchase,
        main = "Boxplot of Purchases",
        xlab = "Purchase",
        ylab = "Amount Spent",
        col = "lightblue",  
        border = "darkblue") 

#dev.off()
```

Using histogram we can clearly see whether the variable are normally
distributed. From the figure below we have 3 histograms for the
variable\
Histogram to see the distribution of the data-set

```{r,  echo=FALSE,cache.comments=TRUE,fig.height= 3,fig.width=8,out.width='100%', fig.cap="Histogram for Age,Puchase and Current City"}
#pdf("hist_combined.pdf", width = 12, height = 5)
par(mfrow = c(1,3))
hist(num_data$Purchase, probability = TRUE,breaks = 30,
     main = "Histogram for Purchase ", col="maroon")
lines(density(num_data$Purchase), col = "green", lwd = 2, lty = 2)
# adding a vertical line
abline(v= mean(num_data$Purchase), col = "purple", lty = 2 , lwd = 2 )

#Legend
legend("topleft", lty = c(2,2),
       col = c("green", "purple"), legend = c("Density", "Average"))


hist(num_data$Age_num, probability = TRUE,breaks = 30,
     main = "Histogram for Age Num ", col="pink")
lines(density(num_data$Age_num), col = "red", lwd = 2, lty = 2)
# adding a vertical line
abline(v= mean(num_data$Age_num), col = "blue", lty = 2 , lwd = 2 )

#Legend
legend("topleft", lty = c(2,2),
       col = c("pink", "blue"), legend = c("Density", "Average"))

hist(num_data$Stay_In_Current_City_Years, probability = TRUE,breaks = 30,
     main = "Histogram forCurrent_City_Years ", col="yellow")
lines(density(num_data$Stay_In_Current_City_Years), col = "grey", lwd = 2, lty = 2)
# adding a vertical line
abline(v= mean(num_data$Stay_In_Current_City_Years), col = "brown", lty = 2 , lwd = 2 )

#Legend
legend("topleft", lty = c(2,2),
       col = c("green", "purple"), legend = c("Density", "Average"))

#dev.off()
```

For the descriptive statistics for categorical variables contingency
tables were used to check the association of the variable together we
bar plots to distinguish the relationships. Below is a snippet of the
categorical data-set.

```{r, echo=FALSE,cache.comments=TRUE,comment= NA,include=FALSE}
cat_data <- data[, c("Gender", "City_Category","Marital_Status")]
head(cat_data,3)
```

Contingency table below show the association between gender,City
category and Marital status

```{r, echo=FALSE,cache.comments=TRUE,fig.height= 3,fig.width=8,warning=FALSE}
contingency <- table(cat_data$Gender, cat_data$City_Category, cat_data$Marital_Status)

contingency_df <- as.data.frame(as.table(contingency))
contingency_df$Percentage <- round((contingency_df$Freq / sum(contingency_df$Freq)) * 100, 2)

library(knitr)
kable(contingency_df, 
      col.names = c("Gender", "City Category", "Marital Status", "Frequency", "Percentage (%)"),
      caption = "Three-variable Contingency Table with Percentages")

```

Mosaic plot If the Male section has wider City Category = A rectangles
compared to City Category = B, it suggests that males are more common in
City Category = A. If the Single slices are taller than Married slices
in all sections, it indicates more singles across all combinations.

```{r mosaic-plot, echo=FALSE, fig.height=3, fig.width=8, out.width='100%', fig.cap="Mosaic Plot of Gender, City Categories, and Marital Status"}
#pdf("plots.pdf", width = 12, height = 5)
par(mfrow =c(1,3))
contingency <- table(cat_data$Gender, cat_data$City_Category, cat_data$Marital_Status)
mosaicplot(contingency, main = "Mosaic Plot", color = TRUE)
cont <- table(cat_data$Gender, cat_data$City_Category)
#round((prop.table(contingency)*100),2)


barplot(cont, beside = TRUE, col = c("blue", "orange"), 
        legend = TRUE,
        main = "Barplot for Gender and City_Category", 
        xlab = "City_Category", 
        ylab = "Frequency")

cont_table <- table(cat_data$Marital_Status, cat_data$Gender)
#round((prop.table(cont_table)*100),2)

barplot(cont_table, beside = TRUE, col = c("pink", "red"),
        legend = TRUE,
        main = "Barplot for Gender and Marital Status", 
        xlab = "Gender",
        ylab = "Frequency")


#dev.off()
```

Simple Linear Regression Model

```{r, echo=FALSE,cache.comments=TRUE,comment= NA,include=FALSE}
smodel <- lm(data$Purchase~data$Age_num)
smodel
```

```{r, echo=FALSE,cache.comments=TRUE,comment= NA, include=FALSE}
summary(smodel)
```

```{r,echo=FALSE,cache.comments=TRUE,comment= NA,include=FALSE}
vcov(smodel)
```

2 (a) Fit the model by estimating the intercept, the slope, and the
underlying uncertainties. (b) Interpret your model. (c) Comment on the
usefulness of your model. (d) Comment on the limitations of your model.
(e) Suggest two ways to improve your model.

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
x <- data$Age_num
y <- data$Purchase
x_bar <- mean(x)
y_bar <- mean(y)
beta_1 <- sum((x- x_bar)*y)/sum((x-x_bar)^2)
beta_1
beta_0 <- y_bar - beta_1*x_bar
beta_0
```

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
n <- length(data$Gender)
n
```

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
fitted_mod <- beta_0 + beta_1*x
head(fitted_mod)
```

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
res <- (y - fitted_mod)
head(res)
```

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
sigma <- sum((y -fitted_mod)^2)/ (n-2)
sigma
```

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
var_beta_0 <- sum(x^2)/(sum((x-x_bar)^2)*n)
var_beta_0
```

```{r, include= FALSE,echo=FALSE,cache.comments=TRUE,comment= NA}
var_beta_1 <- sigma/(sum((x-x_bar)^2))
var_beta_1
```

Model Checking of the Assumptions - Linearity - Homoskedasticity -
Normality

```{r, echo=FALSE,cache.comments=TRUE,comment = NA,fig.height= 3 , fig.width=8, out.width='100%', fig.cap="Plots for Model diagnosis"}
#pdf("model.pdf",width = 12, height = 5)
par(mfrow = c(1,3))

# Check for the assumption of Linearity and homoscendasticity
plot(smodel$residuals ~ smodel$fitted.values)

# Check for the Normality
qqline(smodel$residuals)
qqnorm(smodel$residuals)

#Outlier
plot(smodel, which =4)

#dev.off()
```
