---
title: "Rstudio Introduction"
author: "Moreen Mbuki"
date: "2024-09-13"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
setwd("C:/Users/Administrator/Desktop/AIMS/AIMS SKILL PHASE/IntroCL/Rstudio")
```

```{r}

```


```{r}
# Addition 
1+1
```

# Subtraction
```{r}
2-1
```

# Multipliaction

```{r}
3*3
```

# Division
```{r}
14/5
```

 # Exponent 
```{r}

2**10 ; 2^10
```


# Modulus (Remainder from Division)

```{r}
134 %% 3
```

# Generate numbers and the do the modulus
```{r}
(1:10) %% 2
3 %% 2 #You should get 1 because 3 is not a multiple of 2 
 
```

#  Integer Division

```{r}
5 %/% 2 
```

# Logical Operations 
# Less than 
```{r}
2<2 
```


# Less than or equal to 
```{r}
2<=2 
```


# Greater than 
```{r}
2>4
```
 

# Greater than or equal to 
```{r}
4>= 6
```


# Negation 
```{r}
!TRUE # or !T
!(4>= 6)

```

# AND & used in if statements 
```{r}
TRUE & TRUE; TRUE & FALSE 
T & T # TRUE
T & F # FALSE
F & F # FALSE  

```


#  OR | used on if statement 
```{r}
TRUE | TRUE; TRUE |FALSE
T | T # TRUE
T | F # True
F | F # FALSE  

T & T 
```


## R OBJECTS  To assign a value on R we used <- , = or ->

```{r}
a <- 1 # or 1-> a
x = 2 # Mostly used in data frames
100 -> y 

```

# Vector ,set of value or string 
```{r}
vect <- c("a"= 1 , "b" =2)
vect

```

# To resign 
```{r}
 x=1
 y=x

```

#  VARIABLE IN R 
```{r}
PI <- pi 
PI
class(PI) # To check the class
typeof(PI)
```


# Type of variables
```{r}
n <- 55L;class(n);typeof(n) # Integer

is.integer(n)
n <- 55;class(n);typeof(n) # Numeric 
is.numeric(n)

```

# Complex Numbers 
```{r}
z <- 9 + 3i; class(z) ; typeof(z)
```


# Boolean 
```{r}
true <- TRUE ; class(true)
```


# Character
```{r}
string <- "This is a string" ; class(string) ; is.character(string)

string == tolower("This is a string")
```


#  DATA STRUCTURE AND DATA TYPES 
#  Scalar - set of natural, real, complex, rational numbers
# Vector - collection of identical objects of the same type ; can be a sequence
```{r}
v <- c(1, "R", T, FALSE, NA)
v ; class(v) ; typeof(v)
print(v)
class(NA)
class(1)

v1 <- c(1, 4, T, FALSE, NA)
class(v1)
print(v1)

```


# Sequence 
```{r}
x1 <- seq(0, 2*pi , length.out = 50)
x1 <- round(x1,2)
signif(x1)

x1

length(x1)
x <- seq(0, 2*5 , by =0.1282283)
length(x)
diff(x) # common difference
unique(diff(x))

```


# Replication
```{r}
rep(1,10)
rep(1, 100000) # number to replicate and how many times 
rep(c(0,1), 50)
rep(c(0,1),c(50,50))
rep("String", 50)

trunc(x) # Gives the whole numbers only
ceiling(x)

```

# Sampling 
# dice
```{r}
sample(6,size = 1)

sample(6,size = 1, prob = rep(1/6,6))

smpl <- sample(c("Male","Female"), size = 100, replace = TRUE, prob = rep(1/2,2))

smpl <- sample(c("Male","Female"), size = 100, replace = TRUE, prob = c(30,70)/100)
```


# How many males and female do we expect to have

```{r}
tbl <- table(smpl)
prop.table(table(tbl))

sample(0:1, size = 100, replace = TRUE, prob = c(0.5,0.5)) -> y
y
sum(y == 1) ; sum(!y == 1); sum(y != 1)

```



# Stimulation 
#  Setting the seed for Replication
```{r}
set.seed(11092024)
x <- sample(c("Male", "Female"), size = 10, replace = TRUE)
x

```


####################### R DAY 2 ##############################################
# R Data types / Structure 
# Types example - Scalars, vectors, list, arrays and data frame 


```{r}
table(x) # most applies on factor = binary , categorical , nominal, ordered 
prop.table(table(x))
```


#  confusion matrix#

```{r}
set.seed(12092024) # used to fix randomness  
age <- sample(0:120, size = 100)
(brks <- seq(0,120 , by = 10))

age_groups <-cut(age, breaks=brks, include.lowest = TRUE, right =TRUE ) 
table(age_groups)
```

# To check the missing values
```{r}
age_groups == 0
(which(is.na(age_groups)) -> id_missing)
sum(table(age_groups)) # To check the count
age_factor <- factor(age_groups)
identical(age_factor,age_groups)

```


# Subsetting 
```{r}
age[age >= 80]

freq <- table(age_groups)
```

#Bar plot 
```{r}

barplot(freq) 
```

## Use horizon = TRUE 
```{r}
barplot(freq, horiz = TRUE) 

sum(age >= 80)
mean(age >= 80) * 100 # gives probability 
scatter.smooth(age)
scatter.smooth(freq)
```

#  Pie plot 
```{r}
pie(freq)

```

```{r}
set.seed(12092024)
x<- sample(1:120, size = 1000, replace = TRUE);hist(x)
hist(x, probability = TRUE)
hist(x, )
```

# Density Plot 
```{r}
plot(density(x))
```



################################# R Day 3 ##################################### ######## DATA STIMULATION (Normal, Poisson, Exponential, Gamma and Beta) ###### ###############################################################################

#  Uniform Distribution 
```{r}
unif_dist <- runif(1000) 
```

# Histogram 
```{r, echo=TRUE, fig.height=4 , fig.width=5 }
set.seed(13092024)
unif_dist <- runif(1000) 
hist(unif_dist, probability = TRUE,
     main = "This is the title", col="pink")
lines(density(unif_dist), col = "red", lwd = 2, lty = 2)
# adding a vertical line
abline(v= mean(unif_dist), col = "blue", lty = 2 , lwd = 2 )

#Legend
legend("topleft", lty = c(2,2),
       col = c("pink", "blue"), legend = c("Density", "Average"))
```
## This is my section 
### This is a subsection
#### This is asubsubsection


```{r echo=TRUE,fig.height=4 , fig.width=5 }
rbinom_dist <- rbinom(10000, 10, 0.5)
table(rbinom_dist)
hist(rbinom_dist, probability = TRUE, main = "This is the title", col="pink")
lines(density(rbinom_dist), col = "red", lwd = 2, lty = 2)
# adding a vertical line
abline(v= mean(rbinom_dist), col = "blue", lty = 2 , lwd = 2 )

#Legend
legend("topleft", lty = c(2,2),
       col = c("pink", "blue"), legend = c("Density", "Average"))
```
```{r echo=TRUE,fig.height=4 , fig.width=5}
gauss_dist <- rnorm(1000, mean = 0, sd= 1)
hist(gauss_dist, probability = TRUE,breaks = 30,
     main = "Histogram for Stabdard normal dist", col="pink")
lines(density(gauss_dist), col = "red", lwd = 2, lty = 2)
# adding a vertical line
abline(v= mean(gauss_dist), col = "blue", lty = 2 , lwd = 2 )

#Legend
legend("topleft", lty = c(2,2),
       col = c("pink", "blue"), legend = c("Density", "Average"))
```


# Poisson Distribution 
```{r}
rpois(100,4)
```

```{r echo=TRUE, figo5,fig.height=4 , fig.width=10}
set.seed(13092024)
x <- rnorm(1000)
y <- rnorm(1000, mean = 5, sd = 1.5)
z <- 4 + 8*x +rnorm(1000) # Linear dependent between x and z 

# My Scatter plot
par(mfrow = c(1,2))
plot(y ~ x , main = "Scatter Plot for x and y ",col = "blue", pch = 16)
plot(z ~ x , main = "Scatter Plot for x and y ",col = "blue", pch = 16)
```
#Simulation 
# 1) Standard normal distribution of size n = 1000
#  - histogram 
#  - overlay it density function
# Binormial 
# Poisson 
# Exponential 


# Standard Normal Distribution
```{r}
set.seed(16092024)
stand_dist <- rnorm(100,mean = 0,sd = 1)
hist(stand_dist,probability = TRUE,xlab = NULL, 
     main =" Histogram  of standard naormal distribution", col = "turquoise")
lines(density(stand_dist),col = "red", lwd = 2, lty = 2)
abline(v = mean(stand_dist), col = "blue", lty = 2 , lwd = 2)
legend("toplef", lty = c(2,2),
       col = c("red","blue"), 
       legend = c("Density","Average"))



```


```{r}
set.seed(16092024)
binom_dist <- rbinom(10,2,0.5)
hist(binom_dist,probability = TRUE,xlab = NULL, 
     main =" Histogram  of Binormal distribution", col = "purple")
lines(density(binom_dist),col = "pink", lwd = 2, lty = 2)
abline(v = mean(binom_dist), col = "green", lty = 2 , lwd = 2)
legend("toplef", lty = c(2,2),
       col = c("pink","green"), 
       legend = c("Density","Average"))
```


```{r}
set.seed(16092024)
pois_dist <- rpois(1000,50)
hist(pois_dist,probability = TRUE,xlab = NULL, 
     main =" Histogram  of Poision distribution", col = "black")
lines(density(pois_dist ),col = "yellow", lwd = 2, lty = 2)
abline(v = mean(pois_dist ), col = "brown", lty = 2 , lwd = 2)
legend("toplef", lty = c(2,2),
       col = c("yellow","brown"), 
       legend = c("Density","Average"))
```
# Exponential distribution
```{r}
set.seed(16092024)
Expon_dist <- rexp(1000, rate = 1)
hist(Expon_dist,probability = TRUE,xlab = NULL, 
     main =" Histogram  of Exponential distribution", col = "grey")
lines(density(Expon_dist),col = "maroon", lwd = 2, lty = 2)
abline(v = mean(Expon_dist ), col = "orange", lty = 2 , lwd = 2)
legend("toplef", lty = c(2,2),
       col = c("maroon","orange"), 
       legend = c("Density","Average"))
```

# Gamma distribution
```{r}
set.seed(16092024)
Gamma_dist <- rgamma(100, 1,0.75)
hist(Gamma_dist,probability = TRUE,xlab = NULL, 
     main =" Histogram  of Gamma distribution", col = "wheat4")
lines(density(Gamma_dist),col = "springgreen", lwd = 2, lty = 2)
abline(v = mean(Gamma_dist ), col = "tan3", lty = 2 , lwd = 2)
legend("toplef", lty = c(2,2),
       col = c("springgreen","tan3"), 
       legend = c("Density","Average"))
```


```{r}
set.seed(16092024)
Student_dist <- rt(100, 0.9,37.62)
hist(Student_dist,probability = TRUE,xlab = NULL, 
     main =" Histogram  of Gamma distribution", col = "wheat4")
lines(density(Student_dist),col = "springgreen", lwd = 2, lty = 2)
abline(v = mean(Student_dist ), col = "tan3", lty = 2 , lwd = 2)
legend("toplef", lty = c(2,2),
       col = c("springgreen","tan3"), 
       legend = c("Density","Average"))
```
# Matrice 
```{r}
(A <- matrix(1:25, nrow = 5, ncol = 5))
```

# () used to assign and print
```{r, echo=TRUE}
(B <- matrix(1:25, nrow = 5, ncol = 5, byrow = T)) # ncol is optional especially for a square matrix
```

```{r}
(C <- matrix(1:25, ncol = 5, byrow = F))
```

```{r}
(D <-matrix(c(1,2,3,4,5,6,7,8), nrow = 2))

```

```{r}
rnorm(25,0,1)
```

```{r}
(F <-matrix(rnorm(25,0,1), nrow = 5, ncol = 5))
```
#Tranpose 
# Addition 
# Subtraction
# Multiplication by scalar 
# Point -wise Multiplication
# Trace

#Tranpose
```{r}
(A <- matrix(1:25, nrow = 5, ncol = 5))
(B <- matrix(1:25, nrow = 5, ncol = 5, byrow = T))
#Transpose 
(t(A))

# Addition 
(A + B)

#Solve() get the inversible


```
```{r}
v1 <- c(1,2,3,5)
v2 <- c(4,5,6,8)
v3 <- c(4,6,7,8)
  
(M <- cbind(v1,v2,v3))
(N  <- rbind(v1,v2,v3))


```

```{r}
M <- c(1,2,3,4,5,6,7,8,9)
dim(M) <- c(3,3)
dim(M)
M
```
```{r}
D <- matrix(c(2,2,1,-1), nrow =2, byrow = T)
h <- c(0,2)
solve(D,h)
```

```{r}
M <- c(1,2,3,4,5,6,7,8,9)
M[1:5]
(M[(seq (1,length(M), by=2))])

```
# Multiplication
```{r}
(A <- matrix(1:25, nrow = 5, ncol = 5))
(B <- matrix(1:25, nrow = 5, ncol = 5, byrow = T))

dim(A)[2] == ncol(A)
ncol(A) == nrow(B)
A %*% B
D <- matrix(c(1,2,4,2),nrow = 2,byrow = T)
E <- (solve(D))
D %*% E

```
# Rstudio 17/9/2024
# ARRAY 
#

# Example od Climate data   log1, log2, ---- , logp 
# 
```{r}
(aRray <- array(1:24, dim = c(2,3,2)))
```


```{r}
class(aRray)
```
```{r}
 dim(aRray)
```
#An array can have more than 2 dimension
# It can be of 1 dim (vector), 2 dim (matrix), 3dim the data in ncdf - Climate 

#Matrix in array 
```{r}
 (aRray <- array(1:24, dim = c(2,3,1)))
```
 
```{r}
aRray[2,3,1]

```
# Accessing in Arrays
```{r}
aRray[1,1,1]
```

#
```{r}
aRray[,,1]
```


```{r}
aRray[1,,1]
```
```{r}
airquality
```


```{r}
(aRray <- array(1:24, dim = c(3,4,2)))
```

```{r}
aRray[3,,]
```

```{r}
aRray[,,]
```

```{r}
aRray[,2,]
```

```{r}
x <- rnorm(1000,0,1)
mylist <- list("matrix " = A, "sequence" =head(x), "bool" =TRUE, "Array"= aRray )
mylist
```

```{r}
mylist[[1]]
mylist$`sequence`
mylist[["Array"]]
```

```{r}
mylist[["bool"]] = FALSE
mylist$bool
```

```{r}
mylist[["bool"]] = NULL #Delete the element from the list 
mylist$bool 
mylist
```

```{r}
getwd()
```
```{r}
setwd("C:/Users/Administrator/Desktop/AIMS/AIMS SKILL PHASE/IntroCL/Rstudio")
```
```{r}
getwd()
```

```{r}
setwd("C:/Users/Administrator/Desktop/AIMS/AIMS SKILL PHASE/IntroCL/Rstudio")
```


```{r}
getwd()
```

```{r}
setwd("C:/Users/Administrator/Desktop/AIMS/AIMS SKILL PHASE/IntroCL/Rstudio")
```
```{r}
getwd()
```

```{r}
dir.create( ".\ datafile/")
```
```{r}
# to create a folder in Rstudio

# to get list of all the documents in the directory
dir("C:/Users/Administrator/Desktop/AIMS/AIMS SKILL PHASE/IntroCL/Rstudio")
```

```{r}
# To access csv files in the list
file_names <- dir("./datafile/", pattern = ".csv")
file_names 

List <- list()

List[[file_names[[1]]]]
List["x_1.csv"] <- read.csv("./datafile/x_1.csv")
head(List["x_1.csv"])


# For LOOP
List <- list()
for (l in file_names){
  List[l] <- read.csv(paste0("./datafile/",l))
  
}
print(List)
class(List)
```


#vector 
```{r}

data_l =list()
for (x in file_names) {
  fp <- paste0("./datafile/",x)
  data_l[x] <- read.csv(fp)
  
}
print(data_l)
class(print(data_l))
```

#DATA FRAME 
```{r}
data_l =list()
for (x in file_names) {
  fp <- paste0("./datafile/",x)
    data_l[[x]] <- read.csv(fp)
  
}
print(data_l)
class(data_l)
```

################################ Rstudio Day 5 ################################
###############################################################################


# FOR LOOP
# i take the number in l
```{r}
l <- c(1,2,3,4,5,6,8,9,10)
for(i in l){
  print(i)
}
```


```{r}
dir(pattern =".csv")

```

```{r}
fn <- dir("./datafile/")
fn
```

```{r}

fn <- dir("./datafile/")
fn
path <- "./datafile/"
L <- list()

for (u in fn) {
  #print(urwana)
  L[u] <- read.csv(paste0(path,u))
 }
L
```
```{r}
 setwd("C:/Users/Administrator/Desktop/AIMS/AIMS SKILL PHASE/IntroCL/Rstudio")
```


```{r}

fn <- dir("./datafile/")
fn
path <- "./datafile/"
L <- list()

for (u in fn) {
  #print(urwana)
  L[u] <- read.csv(paste0(path,u))
 }
L
```

```{r}
string1 <- "my"
string2 <- "name is Moreen"
paste0(string2,string1)
```

```{r}
fn <- dir("./datafile/")
fn
path <- "./datafile/"
L <- list()

for (u in fn) {
  #print(urwana)
  L[u] <- read.csv(paste0(path,u))
 }
L

length(L)

# Covert list into df 
my_df <- data.frame(L)
my_df

# renaming column the file 

newnames <- gsub(pattern = ".csv|_", replacement = "", x = colnames(my_df))
#newnames <- gsub(pattern = "_", replacement = "", x = newnames)
#newnames <- gsub(pattern = "[0-9]", replacement = " ", x = newnames)
colnames(my_df) <- newnames
head(my_df)

X10 <- my_df$x10 
my_df2 <- my_df[-2]
my_df2$x10 <- X10
my_df2$x11 <- cos(my_df2$x2)
my_df2
```

```{r}
plot(x11 ~ x2, data = my_df2, col= "green", pch = 8)
```


```{r}
par(mfrow=c(1,2))
hist(my_df2$x11)
p <- "My new name"
with(my_df2, hist(x11, xlab = p, main= paste("Histogram of", p)))


for( i in 1:ncol(my_df2)){
  my_varname <- colnames(my_df2)[i]
  hist(my_df2[,i], main = paste("Histogran of " ,p ), xlab = p)
}
```


```{r}
is.list(my_df2)
is.data.frame(L)
```

```{r}
L2 <- as.list(my_df2)
L2

L3 <- list(my_df)
L3
```

#Deal with missing values 
```{r}
x <- c(NA,1,2,NA,4,NA,NA)
is.na(x)
which(is.na(x)) # which of the elements of x are missing 
which(x == 2)
which(x >= 2)  # which of the element are greater than 2 
which(!is.na(x)) # Which of the elements of x are not missing 


mis_id <- which(is.na(x))
x[mis_id]
x[is.na(x)] <- 9
x[is.na(x)] <- mean(x,na.rm = TRUE)
print(x)
x[is.na(x)] <- mean(x[which(!is.na(x))])
print(x)
y <- x[is.na(x)] <- mean(x,na.rm = TRUE)
print(x)
print(y)

x[is.na(x)] <- mean(x[which(!is.na(x))])
```
```{r}
x <- c(0:10, 50)
xm <- mean(x)
c(xm, mean(x, trim = 0.10))
```

```{r}
x<- c(2,1,2,7,3,2.5,9,"2,7")
class(x)
#converting into numeric 
z<-as.numeric(x)
z
x[is.na(z)]
x[which(is.na(z))] <- 2.7
x
x <- as.numeric(x)
x

```

#################################### 19/9/2024 #############################################
```{r}
head(dat <- datasets::attitude)
# remove a column 
class(dat)
dim(dat)
a <- dat[-1,]
dim(a)
```

############################## While loop in Rstudio  ######################################

```{r}
while (condition) {
  ## code to ne exceuted 
  
  
  ## increment 
  
}
```

# While loop for even numbers 
```{r}
#Declare i 
i <- 1 
while (i <=10) {
  ## code to ne exceuted 
  print(i *2)
  
  ## increment 
  i<- i +1
}
```

# While loop for even numbers 
```{r}
#Declare i 
i <- 1 
while (i <=10) {
  ## code to be executed 
  print(i *2)
  
  ## increment 
  i<- i +10
}

```


```{r}

if(conditions){
  #The code goes here 
}else {
  # codes
} 
```


#If else statements 
```{r}
if(conditions){
  #The code goes here 
}else if (conditions){
  # codes
} else if (conditions){
  #Codes 
  
} else {
  # codes 
}

```


#Exercises 
# Write a program that will tell the User "YOU WON" 

```{r}
#num <- as.integer(readline("Enter a number:"))
num <- 0
while (TRUE) {
  sampe <- sample(6, size =1)
  cat("You got",sampe,"\n")
  print(sampe)
  if (sampe == 6){
    num <- num + 1
    if (num == 3){
      cat("You Won","\n ")
      break 
      }
    } else {
    cat("continue")
  }
}
#print("You won!")
```
#TO get even number only 
```{r}
#print("Enter a number:")
num <- 0
while (num <= 2) {
  sampe <- sample(6, size =1)
  print(sampe)
  if (sampe %% 2 == 0){
    num <- num + 1
  }

}
print("You won!")
```

```{r}
#print("Enter a number:")
num <- 0
while (num <= 2) {
  sampe <- sample(6, size =1)
  print(sampe)
  if (sampe == 1){
    num <- num + 1
  }

}
print("You won!")
```


# For the multiple of 3
```{r}
#print("Enter a number:")
num <- 0
while (num <= 3) {
  sampe <- sample(6, size =1)
  print(sampe)
  if (sampe %% 3 == 0){
    num <- num + 1
  }

}
print("You won!")
```

# REPEAT

# Natural Numbers 
```{r}
i <- 0
repeat{
  print(i)
  i <- i + 1 
  if (i >10) break
}
```
```{r}
i <- 0
repeat{
  print(i)
  i <- i + 2 
  if (i >10) break
}
```

```{r}
num <- 0
repeat{
  sampe <- sample(6,1)
  print(sampe)
  if ( sampe == 6) {
    num <- num + 1  
  }
}
```


# apply function 
```{r}
A <- c(1:4)
dim(A)<- c(2,2)
A

?apply
apply(A, MARGIN = 2,sum)
```

```{r}
iris
```

```{r}
head(iris)
apply(iris[-5], MARGIN = 2, mean)

```
```{r}
head(iris)
apply(iris[-5], MARGIN = 1, sd)

```

#can appy on dataset that are tabular 
```{r}

v <- c(1:100)

avg <- function(x){
  sum(x)/length(x)
}

avg(x = v)
 
head(iris)
apply(iris[-5], MARGIN = 2, avg)
apply(iris[-5], MARGIN = 2, median)
apply(iris[-5], MARGIN = 2, mean)
apply(iris[-5], MARGIN = 2, sd)

apply(iris[-5], MARGIN = 2, summary)

```

```{r}
sapply(iris[-5],sum)
sapply(iris, summary)
sapply(v,function(x) x^2)
sapply(v, sum)
```


# can lappy on dataset that are list dataframe vector and gives back a list only 
```{r}
a <-lapply(iris[-5], sum)
write.csv(a,"a.csv") # 

```

```{r}
is.factor(iris$Species)
xx <- factor(iris$Species) # categorize 
tapply(iris$Sepal.Length, iris[5], mean)
```
```{r}
vapply(X=as.list(iris[-5]), quantile , FUN.VALUE = c("0"= 0,"25" =0 ,"50"= 0, "75"=0, "100" = 0)) # renames of the 
```

# Exercise 
# Write a function that takes an "x" as argumebt abd detects that NA the replace then by mean
```{r}
replace_miss <- function(x){
  # HOW TO DETECT
  id_ms <- which(is.na(x))
  x[id_ms] <- mean(x, na.rm = T)
  return(x)
}


```

can lappy on dataset that are list dataframe vector and gives back a list only 



```{r}
replace_miss <- function(x){
  # HOW TO DETECT
  id_ms <- which(is.na(x))
  x[id_ms] <- mean(x, na.rm = T)
  return(x)
}


```














